---
title: Product-maximising partitions
topics: [mathematics]
---

```{r setup, echo = F, message = F, warning = F}
library(dplyr)
library(knitr)
library(scales)

df <- tibble(
  n = c(2:5, 10, 50, 100, 500, 1000),
  q = n %/% 3,
  r = n %% 3,
  solution = ifelse(r == 0, 3 ^ q, ifelse(r == 1, 2 ^ 2 * 3 ^ (q - 1), 2 * 3 ^ q)),
  approx_floor = (n / floor(n / exp(1))) ^ floor(n / exp(1)),
  approx_ceil = (n / ceiling(n / exp(1))) ^ ceiling(n / exp(1)),
  approx = pmax(approx_floor, approx_ceil),
  error = approx / solution
)
```

Let $\newcommand{\N}{\mathbb{N}}\N=\{1,2,\ldots\}$ be the set of positive integers.
A *partition* of $n\in\N$ is a way of writing $n$ as a sum of positive integers, called *parts*.
For example, $1+2+3$ is a partition of $6$, with parts $1$, $2$, and $3$.
Partitions are unique up to rearrangement: $1+2+3$ and $3+2+1$ are the same partition, but $1+2+3$ and $3+3$ are different partitions.

This post discusses the following problem:

> Let $n\ge2$ be a positive integer.
> Find a partition of $n$ whose parts have maximum product.

For example, the parts in $1+2+3$ have product $1\times2\times3=6$, while the parts in $3+3$ have product $3\times3=9$.
Our goal is to find a product-maximising partition for arbitrary $n$.

Let $x_1+x_2+\cdots+x_k$ be a partition of $n$.
If $x_1=1$ then $k\ge2$ (since $n\ge2$) and
$$\begin{align}
\prod_{i=1}^kx_i
&= 1\times x_2\times\prod_{i=3}^kx_i \\
&< (1+x_2)\times\prod_{i=3}^kx_i
\end{align}$$
because the $x_i$ are strictly positive.[^empty-product]
Thus, replacing the partition $x_1+x_2+\cdots+x_k$ with $(1+x_2)+x_3+\cdots+x_k$ delivers a greater product.
Since the $x_i$ can be rearranged arbitrarily, it follows that product-maximising partitions contain no parts equal to one.
Similarly, if $x_1>4$ then
$$\begin{align}
\prod_{i=1}^kx_i
&= x_1\times\prod_{i=2}^kx_i \\
&< 3(x_1-3)\times\prod_{i=2}^kx_i,
\end{align}$$
so we can obtain a greater product by replacing $x_1+x_2+\cdots+x_k$ with $3+(x_1-3)+x_2+\cdots+x_k$.
It follows that product-maximising partitions contain no parts greater than four.
But $2\times2=4$ and $2+2=4$, so we can replace each four with two twos without reducing the parts' product.
Thus, we can obtain a product-maximising partition using only twos and threes.
Finally, if a partition contains three twos then we should replace them with two threes, since $2+2+2=3+3$ but $2^3=8<9=3^2$.

[^empty-product]: If $j>k$ then $\prod_{i=j}^kx_i=1$ [by convention](https://en.wikipedia.org/wiki/Empty_product).

To summarise, we can obtain a product-maximising partition using only twos and threes, with as many threes as possible.
Letting $n=3q+r$ for some $q\in\N\cup\{0\}$ and $r\in\{0,1,2\}$, the maximum product we can obtain is
$$P(n)=\begin{cases}3^q&\text{if}\ r=0\\ 2^2\times3^{q-1}&\text{if}\ r=1\\ 2\times3^q&\text{if}\ r=2.\end{cases}$$
We can approximate this solution by [relaxing](https://en.wikipedia.org/wiki/Relaxation_(approximation)) the integrality constraint on the $x_i$.
For any given $k$, we can find the vector $x^*$ that solves
$$\newcommand{\R}{\mathbb{R}}\max_{x\in\R_+^k}\prod_{i=1}^kx_i\ \text{subject to}\ \sum_{i=1}^kx_i=n \tag{1},$$
where $\R_+$ is the set of positive real numbers.
This vector has $x_i^*=n/k$ for each $i\in\{1,2,\ldots,k\}$, so that $\prod_{i=1}^kx_i^*=(n/k)^k$.[^lagrange]
If there was no integrality constraint on $k$ then we could maximise $(n/k)^k$ by choosing $k=n/e$, where $e\approx2.718$ is Euler's constant.
But $k$ must be an integer, so we should round it to the nearest integer in whatever direction delivers the greatest value of $(n/k)^k$.
Doing so delivers an estimate
$$\hat{P}(n)=\max\left\{\left(\frac{n}{\lfloor n/e\rfloor}\right)^{\lfloor n/e\rfloor},\left(\frac{n}{\lceil n/e\rceil}\right)^{\lceil n/e\rceil}\right\}$$
of $P(n)$, where $x\mapsto\lfloor x\rfloor$ and $x\mapsto\lceil x\rceil$ are the floor ("round down") and ceiling ("round up") functions.

[^lagrange]: One can derive $x_i^*=n/k$ using the [method of Lagrange multipliers](https://en.wikipedia.org/wiki/Lagrange_multiplier).

The table below compares $P(n)$ and $\hat{P}(n)$ for various $n$.
Since $\{2,3\}\subset\mathbb{R}_+$, the partition of $n$ using twos and as many threes as possible is a feasible, but not necessarily optimal, solution to $(1)$.
Thus $P(n)\le\hat{P}(n)$ for each $n\ge2$.
The multiplicative error $\hat{P}(n)/P(n)$ grows exponentially with $n$ because the exponent $k\in\{\lfloor n/e\rfloor,\lceil n/e\rceil\}$ of $(n/k)^k$ grows (increasingly linearly) with $n$, amplifying the error in the approximation $n/k\sim e$ to each part in the partition underlying $P(n)$.

```{r echo = F}
get_scistr <- function(x) {
  p <- floor(log(x, 10))
  ifelse(p <= 3, round(x, 2), sprintf('%.2f&times;10<sup>%d</sup>', x / (10 ^ p), p))
}

df %>%
  select(-approx_floor, -approx_ceil) %>%
  mutate(n = comma(n, accuracy = 1)) %>%
  mutate_at(c('solution', 'approx'), get_scistr) %>%
  select(`$n$` = n, `$P(n)$` = solution, `$\\hat{P}(n)$` = approx, `$\\hat{P}(n)/P(n)$` = error) %>%
  kable(align = 'c', digits = 2)
```

```{r session-info, echo = F}
bldr::save_session_info()
```
