---
title: Social networks in rural India
topics: [culture, networks, software]
summary: |
  I describe IndianVillages, an R package containing data on social networks in rural India.
  These networks are assortatively mixed with respect to caste.
---

```{r setup, echo = F, message = F}
library(dplyr)
library(igraph)
library(IndianVillages)
library(knitr)
library(purrr)
library(tidyr)
```

[IndianVillages](https://github.com/bldavies/IndianVillages) is a new R package containing data on social networks in rural India.
I derived these data from [Banerjee et al.'s (2013)](https://doi.org/10.1126/science.1236498) surveys of households across `r n_distinct(households$village)` [Karnatakan](https://en.wikipedia.org/wiki/Karnataka) villages.
This post describes the derived data and the networks they define.
I also show that the networks are [assortatively mixed](/blog/assortative-mixing/) with respect to [caste](https://en.wikipedia.org/wiki/Caste_system_in_India).

## Data description

IndianVillages provides two tables.
The first, `households`, links each household to its village and caste:

```{r}
library(dplyr)
library(IndianVillages)

head(households)
```

The `hhid` and `village` columns store household and village IDs.
The `caste` column stores caste memberships:

```{r}
count(households, caste, sort = T)
```

Some `caste` values are missing because the surveys were changed during their collection.
About `r round(100 * sum(households$caste == 'OBC', na.rm = T) / sum(!is.na(households$caste)))`% of the households with known castes are in the [Other Backward Class](https://en.wikipedia.org/wiki/Other_Backward_Class) ("OBC").
This exceeds the (disputed) share of OBCs in India's general population during the survey period.

The second table, `household_relationships`, contains information on inter-household relationships:

```{r}
head(household_relationships)
```

The `hhid.x` and `hhid.y` columns store ego and alter household IDs.
The `type` column stores relationship types:

```{r}
count(household_relationships, type, sort = T)
```

These types correspond to questions asked in Banerjee et al.'s surveys.

## Inter-household networks

We can use `households` and `household_relationships` to define social networks among the households in each village.
First, use the `graph_from_data_frame` function from [igraph](https://igraph.org/) to create the network among all households:

```{r}
library(igraph)

net = graph_from_data_frame(
  distinct(household_relationships, hhid.x, hhid.y),
  directed = F,
  vertices = households
)
```

`net` contains `r format(gsize(net), big.mark = ',')` edges: one for each pair of households with at least one social relationship.
There are no between-village relationships in the data, so we can partition `net` into village-specific networks without deleting any edges:

```{r}
library(purrr)

villages = sort(unique(households$village))

village_nets = map(villages, ~subgraph(net, V(net)$village == .))

sum(map_dbl(village_nets, gsize))  # Same as gsize(net)
```

The networks in `village_nets` are too large to describe visually.
Instead, let's compute some of their properties:

```{r}
village_nets_properties = map_df(village_nets, ~{
  comp = components(.)
  giant = subgraph(., comp$membership == which.max(comp$csize))
  tibble(
    Households = gorder(.),
    `Mean degree` = mean(degree(.)),
    `% of households in giant` = 100 * gorder(giant) / gorder(.),
    `Mean distance in giant` = mean_distance(giant)
  )
})
```

I summarize these properties in the table below.
The number of households in each village ranges from `r min(village_nets_properties[[1]])` to `r max(village_nets_properties[[1]])`.
The mean degree of the households in each village ranges from `r round(min(village_nets_properties[[2]]), 2)` to `r round(max(village_nets_properties[[2]]), 2)`.
Most households are in the giant component for their village, and are connected to others in that component via paths of length two or three.

```{r echo = F}
village_nets_properties %>%
  gather(Property, x, factor_key = T) %>%
  group_by(Property) %>%
  summarise(Mean = mean(x),
            `Std. dev.` = sd(x) * (n() - 1) / n(),
            `Min.` = min(x),
            Median = median(x),
            `Max.` = max(x)) %>%
  kable(digits = 2)
```

## Inter-caste mixing

We can use `net` to study the extent of [assortative mixing](/blog/assortative-mixing/) with respect to caste membership.
First, delete the `r format(sum(is.na(households$caste)), big.mark = ',')` households with missing `caste` values:

```{r}
subnet = subgraph(net, !is.na(V(net)$caste))
```

```{r echo = F}
V(subnet)$caste = gsub('Scheduled', 'Sch.', V(subnet)$caste)
```

`subnet` contains `r format(gorder(subnet), big.mark = ',')` households with a mean degree of `r round(mean(degree(subnet)), 2)`.
This is similar to the mean degree in `net`.
The two networks also have similar mean distances between connected households: `r round(mean_distance(subnet), 2)` in `subnet`, versus `r round(mean_distance(net), 2)` in `net`.

Next, compute `subnet`'s mixing matrix:

```{r}
library(bldr)  # https://github.com/bldavies/bldr

mix_mat = get_mixing_matrix(subnet, 'caste')
```

I define `get_mixing_matrix` [here](https://github.com/bldavies/bldr/blob/master/R/get_mixing_matrix.R).
It returns a matrix in which rows and columns correspond to castes, and entries equal the share of edges joining households in each caste pair.
Multiplying these entries by the sum of degrees---which, by the [degree sum formula](https://en.wikipedia.org/wiki/Handshaking_lemma), equals twice the number of edges---yields a table of inter-caste edge counts:

```{r}
mix_mat * (2 * gsize(subnet))
```

For example, `subnet` contains `r format((mix_mat * 2 * gsize(subnet))[1, 3], big.mark = ',')` edges between households in general castes and households in OBC castes.

We can measure the extent of assortative mixing by comparing `mix_mat` to the matrix we'd expect if edges were independent of caste.
This matrix equals the outer product of the row and column sums of `mix_mat`:

```{r}
mix_mat_indep = rowSums(mix_mat) %*% t(colSums(mix_mat))
```

Comparing the traces of `mix_mat` and `mix_mat_indep` allows us to measure mixing overall:

```{r}
tr = function(m) sum(diag(m))

c(tr(mix_mat), tr(mix_mat_indep))
```

So `subnet` contains about twice as many within-caste edges than we'd expect if edges were independent of caste.

We can also compare `mix_mat` and `mix_mat_indep` element-wise to assess which inter-caste relationships are most over-represented:

```{r}
round(mix_mat / mix_mat_indep, 2)
```

So, for example, there are about 51% more OBC-OBC edges than we'd expect if edges were independent of caste, but less than half as many general-OBC edges.

```{r session-info, echo = F}
bldr::save_session_info()
```
