---
title: Birds, Voting, and Russian Interference
tags: [Bird of the Year, voting]
---

```{r setup, echo = F, message = F, warning = F}
knitr::opts_chunk$set(echo = F, message = F, warning = F,
                     dev = 'svg', fig.ext = 'svg',
                     fig.width = 8, fig.height = 4.5)

# Load packages
library(dplyr)
library(ggplot2)
library(knitr)
library(readr)
library(scales)
library(tidyr)

# Set ggplot2 theme
theme_set(
  theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.subtitle = element_text(margin = margin(b = 10), size = 13),
          plot.title = element_text(face = 'bold', margin = margin(b = 10), size = 16),
          strip.text = element_text(face = 'bold', hjust = 0, margin = margin(b = 5), size = 12))
)

# Import data
data <- read_csv('BOTY-votes-2019.csv', col_types = cols(country = 'c'))
```

Since 2005, [Forest and Bird](https://www.forestandbird.org.nz) has run annual elections for New Zealand's [Bird of the Year](https://www.birdoftheyear.org.nz).
This week Radio New Zealand [announced](https://www.rnz.co.nz/news/national/402986/bird-of-the-year-2019-hoiho-takes-the-winning-title) the [yellow-eyed penguin](https://en.wikipedia.org/wiki/Yellow-eyed_penguin) as 2019's winner.
A follow-up [tweet](https://twitter.com/Forest_and_Bird/status/1193720097283567616) by Forest and Bird [raised suspicions](https://www.rnz.co.nz/news/national/403085/bird-of-the-year-russian-interest-in-contest-piques-suspicions-online) about possible Russian interference into the vote's outcome.

Forest and Bird's tweet includes a world map with countries coloured by voter turnout.
The bar chart below presents the same information in a less exciting format.[^data]

```{r countries}
# Generate plot
countries <- c(
  'New Zealand',
  'United Kingdom',
  'Australia',
  'United States',
  'Russian Federation',
  'Germany',
  'Canada',
  'France',
  'Netherlands',
  'Other'
)
plot_data <- data %>%
  count(country) %>%
  mutate(country = ifelse(dense_rank(-n) <= 10, country, 'Other')) %>%
  count(country, wt = n) %>%
  filter(!is.na(country)) %>%
  mutate(country = factor(country, levels = rev(countries)))
plot_data %>%
  ggplot(aes(country, n / 1e3)) +
  geom_col(alpha = 0.33) +
  geom_text(data = filter(plot_data, country == 'New Zealand'), aes(label = country), hjust = 1, nudge_y = -0.1, size = 3) +
  geom_text(data = filter(plot_data, country != 'New Zealand'), aes(label = country), hjust = 0, nudge_y = 0.1, size = 3) +
  coord_flip() +
  labs(x = NULL,
       y = 'Thousands of votes',
       title = 'Votes for Bird of the Year 2019 by country',
       subtitle = 'New Zealand contributed more votes than all other countries combined',
       caption = 'Source: Forest and Bird') +
  scale_x_discrete(expand = c(0, 0), labels = NULL) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank())
  
```

Russian votes account for `r comma(sum(data$country == 'Russian Federation', na.rm = T))` of the `r comma(sum(!is.na(data$country)))` votes with known country of origin.
New Zealand contributed `r comma(sum(data$country == 'New Zealand', na.rm = T))` such votes.
Fully `r comma(sum(is.na(data$country)))` votes had unknown origin and were excluded from the set of votes used to determine the winning bird.

This year's election used an [instant-runoff](https://en.wikipedia.org/wiki/Instant-runoff_voting) system.
Voters reported up to five of their favorite birds, ranked in order of preference.
Beginning with voters' first preferences, birds with the least votes were eliminated sequentially and their votes reallocated to voters' next favorites.
This process continued until one bird remained.

The table below reports the last five birds eliminated by the instant-runoff process among the votes cast from anywhere, from known countries, from New Zealand, from Russia, and from known countries excluding Russia.
The bracketed percentages represent the share of voters from each country who preferred the top two candidates in the final round.
For example, 61.6% of New Zealanders with preferences over the yellow-eyed penguin and the kākāpō preferred the former.

```{r cache, eval = F}
# Put votes in tidy format
votes <- data %>%
  mutate(voter = row_number()) %>%
  gather(rank, candidate, vote_1:vote_5) %>%
  mutate(rank = as.integer(gsub('vote_', '', rank))) %>%
  filter(!is.na(candidate)) %>%
  group_by(voter) %>%
  mutate(rank = dense_rank(rank)) %>%
  ungroup() %>%
  arrange(voter, rank)

# Define function for computing results of instant runoff vote
get_results <- function(votes) {
  
  # Initialise results
  n_candidates <- n_distinct(votes$candidate)
  res <- tibble(
    candidate = rep('', n_candidates),
    losing_votes = rep(0, n_candidates),
    possible_votes = rep(0, n_candidates)
  )
  
  # Iteratively allocate last place candidates to next preferred
  remaining_votes <- votes
  for (i in seq_len(nrow(res) - 1)) {
    counts <- remaining_votes %>%
      filter(rank == 1) %>%
      count(candidate)
    res$candidate[i] <- slice(counts, which.min(n))$candidate
    res$losing_votes[i] <- min(counts$n)
    res$possible_votes[i] <- sum(counts$n)
    remaining_votes <- remaining_votes %>%
      filter(candidate != res$candidate[i]) %>%
      group_by(voter) %>%
      mutate(rank = dense_rank(rank)) %>%
      ungroup()
  }
  
  # Identify winner
  res$candidate[n_candidates] <- remaining_votes$candidate[1]
  res$losing_votes[n_candidates] <- nrow(remaining_votes)
  res$possible_votes[n_candidates] <- nrow(remaining_votes)
  
  # Output results
  res
}

# Compute results
results_tot <- get_results(votes)
results_known <- get_results(filter(votes, !is.na(country)))
results_nzl <- get_results(filter(votes, country == 'New Zealand'))
results_rus <- get_results(filter(votes, country == 'Russian Federation'))
results_norus <- get_results(filter(votes, country != 'Russian Federation' & !is.na(country)))

# Collate data
table_data <- bind_rows(
  mutate(results_tot, source = 'All countries'),
  mutate(results_known, source = 'Known countries'),
  mutate(results_nzl, source = 'New Zealand'),
  mutate(results_rus, source = 'Russia'),
  mutate(results_norus, source = 'Known countries ex. Russia')
)

# Save cache
write_csv(table_data, 'cache.csv')
```

```{r table}
# Load cache
table_data <- read_csv('cache.csv')

# Generate table
table_data %>%
  group_by(source) %>%
  mutate(rank = dense_rank(-losing_votes)) %>%
  arrange(rank) %>%
  # Compute preferences in final round
  mutate(share = case_when(rank == 2 ~ losing_votes / possible_votes,
                           rank == 1 ~ 1 - (lead(losing_votes) / lead(possible_votes)),
                           T ~ 0)) %>%
  ungroup() %>%
  filter(rank <= 5) %>%
  # Apply formatting
  mutate(candidate = paste0(candidate, ifelse(rank <= 2, paste0(' (', sprintf('%.1f', 100 * share), '%)'), ''))) %>%
  select(Place = rank, source, candidate) %>%
  spread(source, candidate) %>%
  select(c(1, 2, 3, 5, 6, 4)) %>%
  kable(align = 'c')
```

Excluding votes from unknown countries did not affect which bird won.
New Zealand voters got the outcome for which they voted, whereas Russian voters would have crowned the kākāpō.
Removing Russian votes wouldn't have changed the election outcome---to the extent that Russians did interfere with the vote, their interference was not successful.

```{r session-info}
options(width = 80)
writeLines(capture.output(sessioninfo::session_info()), 'session.log')
```

[^data]: The data used in this post are copyright Forest and Bird, and are released under a [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/) license. They are available [here](https://www.dragonfly.co.nz/news/2019-11-12-boty.html).
