---
title: Delayed saving
topics: [finance]
hidden: true
---

Suppose I want to retire at time $T>0$.
I make constant payments to a savings account that earns continuously compounded interest $r>0$.
I want my retirement fund to be worth $V>0$ today (time $0$).
How much bigger do my payments have to be if I delay them?

Let $X_d$ be the payments I have to make if I start saving at time $d\in[0,T]$.
These payments form an [annuity](https://en.wikipedia.org/wiki/Annuity) with value
$$\frac{X_d}{r}\left(1-e^{-r(T-d)}\right)$$
at time $d$.
I want this value to equal $Ve^{rd}$.
So my payments must equal
$$\begin{align}
X_d
&= \frac{r}{1-e^{-r(T-d)}}\times Ve^{rd} \\
&= \frac{rV}{e^{-rd}-e^{-rT}}.
\end{align}$$
Therefore, delaying to time $d$ increases my payments by a factor of
$$\frac{X_d}{X_0}=\frac{1-e^{-rT}}{e^{-rd}-e^{-rT}}.$$
The chart below shows how $X_d/X_0$ grows with the proportion of time $d/T$ I delay saving.
Part of this growth comes from having less time remaining: if my savings earn no interest, then the factor
$$\lim_{r\to0}\frac{X_d}{X_0}=\frac{T}{T-d}$$
equals the ratio of time until retirement and time spent saving.
Raising $r$ raises $X_d/X_0$ because I forgo more opportunities to earn interest on my interest the longer I delay.
This is especially true when I'm far from retiring (i.e., $T$ is large).

```{r setup, echo = F, message = F}
library(bldr)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)

opts_chunk$set(echo = F, message = F,
               fig.width = 6, fig.height = 3, fig.ext = 'svg', dev = 'svg')

set_ggtheme()
```

```{r plot}
crossing(t = c(1, 10, 100), d = seq(0, 1, by = 0.01), r = c(0, 0.01, 0.05)) %>%
  mutate(d = d * t,
         y = ifelse(r == 0, t / (t - d), (1 - exp(-r * t)) / (exp(-r * d) - exp(-r * t)))) %>%
  ggplot(aes(d / t, y)) +
  geom_line(aes(col = paste0('r = ', 100 * r, '%'))) +
  facet_wrap(~paste('T =', t)) +
  labs(x = 'Delay ratio (d/T)',
       y = expression(paste('Payment ratio (', X[d] / X[0], ')')),
       title = 'Delaying payments makes them bigger',
       subtitle = 'Ratio of payments needed to reach a target balance at time T\nwhen I start saving at time d vs. time 0',
       col = NULL) +
  coord_cartesian(expand = F, ylim = c(1, 10)) +
  scale_x_continuous(labels = \(x) round(x, 2)) +
  scale_y_continuous(labels = \(y) round(y, 2)) +
  theme(legend.justification = c(0, 1),
        legend.position = c(0.01, 1),
        panel.spacing.x = unit(1, 'lines'))
```

```{r session-info}
save_session_info()
```

---

*Thanks to [Michael Boskin](https://profiles.stanford.edu/michael-boskin) for inspiring this post.*
