---
title: Correlation and concordance
topics: [mathematics, statistics]
summary: I derive the probability of sampling a concordant pair from the bivariate normal distribution.
---

```{r setup, echo = F, message = F}
library(bldr)
library(dplyr)
library(ggplot2)
library(knitr)
library(MASS)
library(purrr)
library(tidyr)

opts_chunk$set(echo = F, message = F,
               fig.width = 6, fig.height = 3, fig.ext = 'svg', dev = 'svg')

set_ggtheme()
```

Let $X=(X_1,X_2)$ be a random vector in $\mathbb{R}^2$.
Two realizations $x$ and $x'$ of $X$ form a [concordant pair](https://en.wikipedia.org/wiki/Concordant_pair) if $(x_2'-x_2)$ and $(x_1'-x_1)$ have the same sign.
What's the probability of sampling a concordant pair when $X$ is bivariate normal?

```{r scatter_df}
simulate_one = function(rho, n = 1e2) {
  mu = c(0, 0)
  Sigma = matrix(c(1, rho, rho, 1), ncol = 2)
  d = MASS::mvrnorm(n, mu, Sigma)
  return (tibble(x1 = d[, 1], x2 = d[, 2]))
}

get_empirical_cr = function(df) {
  x1 = df$x1
  x2 = df$x2
  n = nrow(df)
  res = 0
  i = 1
  while (i <= n) {
    j = 1
    while (j < i) {
      res = res + ((x2[i] - x2[j]) * (x1[i] - x1[j]) > 0)
      j = j + 1
    }
    i = i + 1
  }
  return (res / (n * (n - 1) / 2))
}

set.seed(0)
scatter_df = tibble(rho = c(-0.5, 0, 0.5)) %>%
  mutate(dat = map(rho, simulate_one),
         rate = map_dbl(dat, get_empirical_cr))
```

For example, suppose $X_1$ and $X_2$ have zero means, unit variances, and a correlation of $\rho$.
The scatter plots below show 100 realizations of $(X_1,X_2)$ when $\rho\in\{-0.5,0,0.5\}$.
These realizations contain
$$\binom{100}{2}=4,\!950$$
pairs, of which `r paste0(round(100 * scatter_df$rate[1]), '%')` are concordant when $\rho=-0.5$.
This percentage rises to `r paste0(round(100 * scatter_df$rate[2]), '%')` when $\rho=0$ and to `r paste0(round(100 * scatter_df$rate[3]), '%')` when $\rho=0.5$.
Increasing $\rho$ makes concordance more likely because it makes $(X_2-X_1)$ larger and less noisy.

```{r scatter}
scatter_df %>%
  unnest('dat') %>%
  ggplot(aes(x1, x2)) +
  geom_point() +
  labs(x = expression(X[1]),
       y = expression(X[2]),
       title = '100 realizations of X',
       subtitle = expression(paste(X[1], ' and ', X[2], ' have zero means, unit variances, and a correlation of \u03c1'))) +
  facet_wrap(~paste('\u03c1 =', rho)) +
  coord_fixed(clip = 'off', expand = F) +
  theme(panel.spacing.x = unit(1, 'line'))
```

Different samples give different concordance rates due to sampling variation.
We can remove this variation by deriving the concordance rate analytically.
To begin, suppose $X$ has mean $\mathrm{E}[X]=(\mu_1,\mu_2)$ and covariance matrix
$$\mathrm{Var}(X)=\begin{bmatrix}
\sigma_1^2 & \rho\sigma_1\sigma_2 \\
\rho\sigma_1\sigma_2 & \sigma_2^2
\end{bmatrix}.$$
Then $X_2\mid X_1$ is normal with mean
$$\mathrm{E}[X_2\mid X_1]=\mu_2+\frac{\rho\sigma_2}{\sigma_1}(X_1-\mu_1)$$
and variance
$$\mathrm{Var}(X_2\mid X_1)=(1-\rho^2)\sigma_2^2.$$
So for any two realizations $x$ and $x'$ of $X$ we can write
$$\renewcommand{\epsilon}{\varepsilon}
x'_2-x_2=\frac{\rho\sigma_2}{\sigma_1}\left(x'_1-x_1\right)+\epsilon$$
with $\epsilon\sim N(0,2(1-\rho^2)\sigma_2^2)$.
Now $x'_1-x_1\sim N(0,2\sigma_1^2)$ is normal, and so
$$z\equiv \frac{x'_1-x_1}{\sigma_1\sqrt{2}}$$
is standard normal and exceeds zero if and only if $x'_1>x_1$.
Letting $f$ and $\phi$ be the density functions for $\epsilon$ and $z$ then gives
$$\newcommand{\der}{\mathrm{d}}
\begin{align}
\Pr(x'_2>x_2\ \text{and}\ x'_1>x_1)
&= \Pr(\sqrt{2}\rho\sigma_2 z+\epsilon>0\ \text{and}\ z>0) \\
&= \int_0^\infty\left(\int_{-\sqrt{2}\rho\sigma_2 z}^\infty f(\epsilon)\,\der \epsilon\right)\phi(z)\,\der z \\
&\overset{\star}{=} \int_0^\infty\left(\int_{\frac{-\rho z}{\sqrt{1-\rho^2}}}^\infty \phi(w)\,\der w\right)\phi(z)\,\der z \\
&= \int_0^\infty\left(1-\Phi\left(\frac{-\rho z}{\sqrt{1-\rho^2}}\right)\right)\phi(z)\,\der z \\
&\overset{\star\star}{=} \frac{1}{2}-\int_0^\infty\Phi\left(\frac{-\rho z}{\sqrt{1-\rho^2}}\right)\phi(z)\,\der z,
\end{align}$$
where $\Phi$ is the standard normal CDF, where $\star$ uses the change of variables
$$w\equiv \frac{\epsilon}{\sigma_2\sqrt{2(1-\rho^2)}},$$
and where $\star\star$ uses the symmetry of $\phi$ about $z=0$.
But $f$ is symmetric about $\epsilon=0$, which implies
$$\Pr(x'_2>x_2\ \text{and}\ x'_1>x_1)=\Pr(x'_2<x_1\ \text{and}\ x'_1<x_1),$$
and therefore
$$\begin{align}
C(\rho)
&\equiv \Pr(x\ \text{and}\ x'\ \text{are concordant}) \\
&= \Pr(x'_2>x_2\ \text{and}\ x'_1>x_1)+\Pr(x'_2<x_1\ \text{and}\ x'_1<x_1) \\
&= 1-2\int_0^\infty\Phi\left(\frac{-\rho z}{\sqrt{1-\rho^2}}\right)\phi(z)\,\der z.
\end{align}$$
The concordance rate $C(\rho)$ depends on the correlation $\rho$ of $X_1$ and $X_2$, but not their means or variances.
It has value $C(0)=0.5$ when $\rho=0$ because $\Phi(0)=0.5$ is constant.
Intuitively, if $X_1$ and $X_2$ are uncorrelated then we can't use $(x'_1-x_1)$ to predict $(x'_2-x_2)$, which is equally likely to be positive or negative.
Whereas if $\lvert\rho\rvert=1$ then $(x'_1-x_1)$ predicts $(x'_2-x_2)$ perfectly, and so
$$\lim_{\rho\to1}C(\rho)=1$$
and
$$\lim_{\rho\to-1}C(\rho)=0.$$
The chart below verifies that the concordance rate $C(\rho)$ grows with $\rho$.
It also shows that
$$C(\rho)+C(1-\rho)=1.$$
Thus, for example, we have $C(-0.5)=1/3$ and $C(0.5)=2/3$.
These values remove the sampling error from the estimates `r round(scatter_df$rate[1], 2)` and `r round(scatter_df$rate[3], 2)` obtained using the 100 realizations above.

```{r integral_df}
get_cr = function(rho) {
  f = \(z) pnorm(-rho * z / sqrt(1 - rho ^ 2)) * dnorm(z)
  return(1 - 2 * integrate(f, 0, 1e2)$value)
}

integral_df = tibble(rho = seq(-1, 1, by = 0.01)) %>%
  mutate(rate = map_dbl(rho, get_cr))
```

```{r integral}
integral_df %>%
  ggplot(aes(rho, rate)) +
  geom_line() +
  labs(x = expression(paste('Correlation \u03c1 of ', X[1], ' and ', X[2])),
       y = 'Concordance rate C(\u03c1)',
       title = 'Concordance rates when X is bivariate normal',
       subtitle = 'C(\u03c1) is increasing in \u03c1 and satisfies C(\u03c1)+C(1-\u03c1)=1') +
  coord_cartesian(clip = 'off', expand = F, xlim = c(-1, 1), ylim = c(0, 1)) +
  scale_x_continuous(labels = \(x) round(x, 2)) +
  scale_y_continuous(labels = \(y) round(y, 2))
```

```{r session-info}
save_session_info()
```
