---
title: Computing Epicycles
topics: [mathematics]
---

```{r setup, echo = FALSE, message = F, warning = F}
knitr::opts_chunk$set(echo = F, message = F, warning = F)

library(bezier)
library(dplyr)
library(gganimate)
library(ggforce)
library(ggplot2)
library(jsonlite)
library(purrr)
library(xml2)
```

Earlier this year Grant Sanderson, creator of the YouTube channel [3blue1brown](https://www.3blue1brown.com), posted a [video](https://www.youtube.com/watch?v=r6sGWTCMz2k) explaining how [Fourier series](http://mathworld.wolfram.com/FourierSeries.html) approximate periodic functions using sums of sines and cosines.
In the video and its [companion](https://www.youtube.com/watch?v=-qgreAUpPwM), Grant animates sets of vectors that rotate on circular orbits and, when summed together, reproduce a range of images defined by closed curves.

Consider, for example, the boundary of GitHub's logo:

```{r plot, dev = 'svg', fig.ext = 'svg', fig.width = 8, fig.height = 4.5}

# Read SVG path as string
path_str <- read_xml('https://simpleicons.org/icons/github.svg') %>%
  xml_children() %>%
  xml_attr('d') %>%
  {.[2]}

# Parse SVG path using svg-path-parser, installed via Node.js:
#
# ```bash
# brew install node
# npm install svg-path-parser
# ```
path_data <- paste0(
  'var d=\'', path_str, '\';',
  'const {parseSVG, makeAbsolute} = require(\'svg-path-parser\');',
  'const commands = parseSVG(d);',
  'makeAbsolute(commands);',
  'console.log(JSON.stringify(commands));'
) %>%
  paste0('node -e \"', ., '\"') %>%
  system(intern = TRUE) %>%
  fromJSON() %>%
  select(-relative)

# Compute coordinates by sampling points on cubic Bezier curves
coordinates <- 2 : nrow(path_data) %>%
  purrr::map(
    function(row, n_points = 20) {
      t <- seq(0, 1, length = n_points)
      p <- c(unlist(path_data[row, 5:10]), unlist(path_data[row, 3:4])) %>%
        matrix(nrow = 4, byrow = TRUE)
      fit <- bezier::bezier(t, p)
      tibble::tibble(x = fit[, 1], y = fit[, 2], row = row)
    }
  ) %>%
  bind_rows() %>%
  mutate(x = x - 12, y = 12 - y) %>%  # Centre at origin
  filter(row == lead(row)) %>%  # Remove duplication at curve endpoints
  select(-row) %>%
  mutate_all(round, 4)

# Plot closed curve
coordinates %>%
  bind_rows(slice(., 1)) %>%  # Close curve
  ggplot(aes(x, y)) +
  geom_polygon(alpha = 0.1) +
  geom_path() +
  coord_fixed(clip = 'off') +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal()
```

Let $\gamma:[0,1]\to\mathbb{R}^2$ be the closed curve in $\mathbb{R}^2$ defining the logo's boundary.
Suppose there is an integer $n$ such that
$$\gamma(t) = \sum_{k=-n}^n \gamma_k(t)$$
for some set of circular orbits $\gamma_{-n},\ldots,\gamma_n:[0,1]\to\mathbb{R}^2$ and for all times $t\in[0,1]$.
(Negative and positive subscripts correspond to clockwise and anti-clockwise orbits.
Both may be necessary to reconstruct $\gamma$.)
Each orbit $\gamma_k$ has time $t$ position defined by the vector
$$\gamma_k(t) = \begin{bmatrix} r_k \cos(2\pi k t + \theta_k) \\ r_k \sin(2\pi k t + \theta_k) \end{bmatrix}$$
for some radius $r_k$, angular speed $2\pi k$ rad/s and initial phase $\theta_k$.
Consequently, the curves $x,y:[0,1]\to\mathbb{R}$ defining the horizontal and vertical components of $\gamma$ must satisfy the system
$$\begin{align} x(t) &= \sum_{k=-n}^n r_k\cos(2\pi k t + \theta_k) \\ y(t) &= \sum_{k=-n}^n r_k\sin(2\pi k t + \theta_k) \end{align}$$
of identities.
Let $z:[0,1]\to\mathbb{C}$ be the curve with $z(t)=x(t)+iy(t)$ for all $t\in[0,1]$.
[Euler's formula](http://mathworld.wolfram.com/EulerFormula.html) gives
$$\begin{align} z(t) &= \sum_{k=-n}^n r_k(\cos(2\pi k t + \theta_k) + i \sin(2\pi k t + \theta_k)) \\ &= \sum_{k=-n}^n r_k \exp(2\pi i k t + i\theta_k) \\ &= \sum_{k=-n}^n c_k \exp(2\pi i k t), \end{align}$$
where each Fourier coefficient $c_k=r_k\exp(i\theta_k)$ has modulus $\lvert c_k\rvert=r_k$ and (principal) argument $\mathrm{Arg}(c_k)=\theta_k$.
Now, notice that
$$\begin{align} \int_0^1 z(t) \exp(-2\pi i k t)\, \mathrm{d}\,t &= \int_0^1\left(\sum_{j=-n}^n c_j \exp(2\pi i j t)\right)\exp(-2\pi i k t)\, \mathrm{d}\,t \\ &= \int_0^1c_k\, \mathrm{d}\,t + \sum_{j\not=k} c_j \int_0^1 \exp(2\pi i (j - k)t)\, \mathrm{d}\,t \\ &= c_k \end{align}$$
for each $k$ because
$$\int_0^1 \exp(2\pi i (j - k)t)\, \mathrm{d}\,t = 0$$
for all integers $j\not=k$ by the $2\pi i$-periodicity of the complex exponential function.
Thus
$$c_k = \int_0^1 z(t) \exp(-2\pi i k t)\, \mathrm{d}\, t,$$
which can be calculated using Riemann sums given sample points along the component curves $x$ and $y$.
Doing this calculation for each $k$, and computing the corresponding moduli $r_{-n},\ldots,r_n$ and arguments $\theta_{-n},\ldots,\theta_n$, provides enough information to generate the animation below.

```{r animation}
# Define function for computing Fourier coefficients for complex signal
get_fourier_coefficients <- function(z, freqs, ts) {
  i <- complex(real = 0, imaginary = 1)
  dt <- ts[2]
  res <- complex(length(freqs))
  for (f in 1 : length(freqs)) {
    # Approximate integral form using Riemann sums
    for (t in 1 : length(ts)) {
      res[f] <- res[f] + exp(-2 * pi * i * freqs[f] * ts[t]) * z[t] * dt
    }
  }
  res
}

# Compute Fourier coefficients
ts <- (0 : (nrow(coordinates) - 1)) / nrow(coordinates)
z <- complex(real = coordinates$x, imaginary = coordinates$y)
frequencies <- -50 : 50
coefficients <- get_fourier_coefficients(z, frequencies, ts)

# Compute vector displacements (relative to orbital centre) at each time steps
displacements <- tibble(
  t = rep(ts, each = length(coefficients)),
  freq = rep(frequencies, length(ts)),
  coeff = rep(coefficients, length(ts))
) %>%
  mutate(radius = Mod(coeff),
         phase = Arg(coeff),
         orbit = dense_rank(abs(freq) + 0.5 * (freq < 0)),
         angle = phase + 2 * pi * freq * t,
         x_disp = radius * cos(angle),
         y_disp = radius * sin(angle))

# Compute vector positions (relative to origin) at each time step
origin_displacement <- tibble(
  t = ts,
  orbit = 0,
  x_disp = 0,
  y_disp = 0
)
positions <- displacements %>%
  select(t, orbit, x_disp, y_disp) %>%
  bind_rows(origin_displacement) %>%
  arrange(t, orbit) %>%
  group_by(t) %>%
  mutate(x_pos = cumsum(x_disp),
         y_pos = cumsum(y_disp)) %>%
  ungroup()

# Compute draw positions at each time step
draw_positions <- positions %>%
  filter(orbit == max(orbit)) %>%
  select(t, x = x_pos, y = y_pos)

# Compute orbital centres and radii
orbits <- positions %>%
  group_by(t) %>%
  mutate(x0 = lag(x_pos),
         y0 = lag(y_pos),
         r = sqrt(x_disp ^ 2 + y_disp ^ 2)) %>%
  ungroup() %>%
  select(t, orbit, x0, y0, r) %>%
  filter(orbit > 0 & orbit < 30)

# Animate epicycles
plot_data <- positions %>%
  left_join(draw_positions) %>%
  right_join(orbits) %>%
  group_by(orbit) %>%
  filter(row_number() %% 3 == 1) %>%  # Skip time steps
  ungroup()
timed_coordinates <- tibble()
for (t in sort(unique(plot_data$t))) {
  timed_coordinates <- timed_coordinates %>%
    bind_rows(mutate(bind_rows(coordinates, slice(coordinates, 1)), t = t))
}
p <- plot_data %>%
  ggplot() +
  geom_circle(data = filter(plot_data, orbit > 1), aes(x0 = x0, y0 = y0, r = r), col = 'grey75') +
  geom_path(aes(x_pos, y_pos), col = 'grey50') +
  geom_polygon(data = timed_coordinates, aes(x, y), alpha = 0.1) +
  geom_point(aes(x, y), size = 4) +
  coord_fixed(clip = 'off') +
  labs(x = NULL,
       y = NULL) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_void() +
  transition_time(t)
animate(p, nframes = n_distinct(plot_data$t), fps = 20, width = 8, height = 4.5, res = 100, units = 'in')
```

```{r session-info, echo = F}
options(width = 80)
writeLines(capture.output(sessioninfo::session_info()), 'session.log')
```
