---
title: "*Armchair Expert* episodes"
topics: [media, software]
---

```{r setup, echo = F, message = F}
library(ArmchairExpert)
library(bldr)
library(dplyr)
library(ggplot2)
library(knitr)
library(lubridate)
library(tidyr)

opts_chunk$set(echo = F, message = F, warning = F,
               fig.width = 6, fig.height = 3, fig.ext = 'svg', dev = 'svg')

set_ggtheme()
```

```{r episodes, cache = T}
episodes = episodes %>%
  mutate(Show = ifelse(is.na(show), 'Interviews', show),
         show_group = ifelse(is.na(show), Show, 'Shows'))
```

[*Armchair Expert*](https://armchairexpertpod.com) is a podcast hosted by [Dax Shepard](https://en.wikipedia.org/wiki/Dax_Shepard) and [Monica Padman](https://en.wikipedia.org/wiki/Monica_Padman).
They interview celebrities, scientists, and other public figures.
They also publish some subsidiary "shows" via the podcast's main feed.
The table below counts episodes by show:

```{r show_counts}
show_counts = episodes %>%
  {bind_rows(., mutate(., Show = 'Total'))} %>%
  count(Show, name = 'Episodes') %>%
  arrange(Show == 'Total', -Episodes)
```

```{r show_counts-table}
kable(show_counts)
```

I store episodes' metadata in the R package [ArmchairExpert](https://github.com/bldavies/ArmchairExpert).
It contains a single table, `episodes`, with a row for each episode and seven columns:

* `id`: Episode ID on Spotify
* `date`: Episode release date
* `title`: Episode title
* `show`: Show to which episode belongs
* `number`: Within-show episode number
* `duration`: Episode length (in seconds)
* `description`: Episode description

<!-- I list episodes' Spotify IDs because I collect their metadata via the [Spotify API](https://developer.spotify.com/documentation/web-api). -->

The first *Armchair Expert* episode---an interview with Dax's wife, [Kristen Bell](https://en.wikipedia.org/wiki/Kristen_Bell)---was released in February 2018.
The earliest show ([Monica & Jess Love Boys](https://armchairexpertpod.com/monica-jess-love-boys)) started in February 2020 and ended two months later.
Other shows have been and gone, and three are ongoing:

```{r show-timelines}
plot_df = episodes %>%
  mutate(Show = factor(Show, rev(show_counts$Show))) %>%
  group_by(Show) %>%
  mutate(start_date = min(date),
         end_date = max(date)) %>%
  slice(1, n()) %>%
  ungroup() %>%
  mutate(ongoing = end_date >= '2024-07-01',
         date = if_else(ongoing & date == end_date, max(end_date), date))

plot_df %>%
  ggplot(aes(date, Show)) +
  geom_line() +
  geom_point(aes(shape = date == start_date, fill = ongoing), show.legend = F) +
  labs(x = 'Date',
       y = NULL,
       title = 'Armchair Expert timeline',
       subtitle = 'Six subsidiary shows have been and gone, and three are ongoing') +
  coord_cartesian(clip = 'off', expand = F) +
  scale_fill_manual(values = c('black', 'white')) +
  scale_shape_manual(values = c(21, 19))
```

My favorite show is [Flightless Bird](https://armchairexpertpod.com/flightless-bird).
It's hosted by [David Farrier](https://en.wikipedia.org/wiki/David_Farrier), a fellow Kiwi who reflects on [living in the USA](/blog/living-america/).

The median episode is about `r median(round(episodes$duration / 60))` minutes long.
But most interviews are longer and most shows are shorter:

```{r lengths}
episodes %>%
  ggplot(aes(duration / 60)) +
  geom_histogram(aes(fill = show_group), alpha = 0.5, binwidth = 5, position = 'identity') +
  labs(x = 'Length (minutes)',
       y = 'Episodes',
       title = 'Armchair Expert episode lengths',
       subtitle = paste('The median episode is about', median(round(episodes$duration / 60)), 'minutes long'),
       fill = NULL) +
  coord_cartesian(clip = 'off', expand = F) +
  scale_x_continuous(breaks = 30 * (0:10)) +
  theme(legend.justification = c(0.01, 1),
        legend.position = c(0.01, 1))
```

Most interviews end with a "fact check," during which Dax and Monica discuss the interview and their lives.
Fact checks can be as long as the interviews themselves.

```{r interviews, cache = T, dependson = 'episodes'}
interviews = episodes %>%
  filter(show_group == 'Interviews') %>%
  select(date, title, number) %>%
  filter(number > 1) %>%
  filter(!grepl('Best of|Holiday Spectacular', title))
```

```{r interview_guests, cache = T, dependson = 'interviews'}
interview_guests = interviews %>%
  mutate(guest = title,
         guest = sub('.*: ', '', guest),
         guest = sub('\\(.*\\)$', '', guest),
         guest = sub('Return.*$', '', guest),
         guest = sub('Men\'s Bodies with', '', guest),  # Kumail Nanjiani and Rob McElhenney
         guest = sub('on COVID-19', '', guest), # Sanjay Gupta
         guest = sub('#4', '', guest),  # Sanjay Gupta and David Sedaris
         guest = sub(', by.*$', '', guest),  # Kristen Bell
         guest = sub('Yuval Harari', 'Yuval Noah Harari', guest)) %>%
  select(guest, number) %>%
  mutate(guest = strsplit(guest, '&')) %>%
  unnest('guest') %>%
  mutate(guest = trimws(guest)) %>%
  left_join(interviews, by = 'number')
```

```{r repeat_guest_lists, cache = T, dependson = 'interview_guests'}
repeat_guest_lists = interview_guests %>%
  count(guest) %>%
  filter(n > 1) %>%
  group_by(n) %>%
  summarise(guest_list = paste(guest, collapse = ', ')) %>%
  ungroup() %>%
  mutate(guest_list = gsub(', ([^,]+)$', ', and \\1', guest_list))
```

Dax and Monica have interviewed some people many times.
They've interviewed Kristen Bell five times, and [David Sedaris](https://en.wikipedia.org/wiki/David_Sedaris) and [Sanjay Gupta](https://en.wikipedia.org/wiki/Sanjay_Gupta) four times each.[^many-interviews]
My favorite interviews are with [Esther Perel](https://en.wikipedia.org/wiki/Esther_Perel), [Wendy Mogel](https://en.wikipedia.org/wiki/Wendy_Mogel), and [Terry Crews](https://en.wikipedia.org/wiki/Terry_Crews).

[^many-interviews]: `r filter(repeat_guest_lists, n == 3)$guest_list` have been interviewed three times each.

```{r session-info}
save_session_info()
```
