---
title: Loan repayments
topics: [finance]
summary: |
  I compare the share of payments made on a loan to the share of the loan repaid.
hidden: yes
---

```{r setup, echo = F, message = F}
library(bldr)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)

opts_chunk$set(echo = F, message = F,
               fig.width = 6, fig.height = 3, fig.ext = 'svg', dev = 'svg')

set_ggtheme()
```

Suppose I take out a loan.
It gains interest at rate $r$, compounded continuously.
I repay the loan by making constant, continuous payments until time $T$.
How does the repaid share of my loan vary over time?
And how does it depend on $r$ and $T$?

Let $P_0$ be the initial value of my loan: the "principal."
Then my continuous payments $C$ must satisfy
$$\begin{align}
P_0
&= \int_0^TCe^{-r\tau}\,\mathrm{d}\tau \\
&= \frac{C}{r}\left(1-e^{-rT}\right)
\end{align}$$
and so the value of my remaining payments at time $t\in[0,T]$ equals
$$\begin{align}
P_t
&\equiv \int_t^TCe^{-r(\tau-t)}\,\mathrm{d}\tau \\
&= \frac{C}{r}\left(1-e^{-r(T-t)}\right) \\
&= P_0\left(\frac{e^{-rt}-e^{-rT}}{1-e^{-rT}}\right)e^{rt}.
\end{align}$$
If I don't make any payments before time $t$ then the principal grows to $P_0e^{rt}$.
Therefore, the value of my repayments up to time $t$ equals the difference $(P_0e^{rt}-P_t)$.

Now let $x\equiv t/T\in[0,1]$ be share of payments I've made up to time $t$.
The chart below plots the corresponding share
$$\frac{P_0e^{rt}-P_t}{P_0e^{rt}}\bigg\rvert_{t=xT}=\frac{1-e^{-xrT}}{1-e^{-rT}}$$
of the loan that I've repaid.
This share grows with $x$ at a decreasing rate.
Intuitively, my repayment "slows down" because the interest on the principal and payments grows larger than the payments themselves.
This slowing effect is stronger when the interest rate $r$ is larger and time horizon $T$ is longer.

```{r plot}
rates = c(0.01, 0.05, 0.1)
horizons = c(10, 30, 100)

crossing(r = rates, TT = horizons, x = seq(0, 1, by = 0.01)) %>%
  mutate(y = (1 - exp(-x * r * TT)) / (1 - exp(-r * TT))) %>%
  ggplot(aes(100 * x, 100 * y)) +
  geom_line(aes(col = factor(TT, horizons, paste('T =', horizons)))) +
  facet_wrap(~factor(r, rates, labels = paste0('r = ', 100 * rates, '%'))) +
  labs(x = '% of payments made',
       y = '% of loan repaid',
       title = 'Loan repayment trajectories',
       subtitle = 'Repayment slows over time, especially when the interest rate r is large\nand the time horizon T is long',
       col = NULL) +
  coord_cartesian(clip = 'off', expand = F) +
  guides(col = guide_legend(label.position = 'left')) +
  scale_color_brewer(palette = 'Set1') +
  theme(legend.justification = c(1, 0),
        legend.position = c(1, 0),
        panel.spacing.x = unit(1, 'lines'),
        plot.margin = margin(5.5, 7.5, 5.5, 7.5, 'pt'))
```

```{r session-info}
save_session_info()
```
