---
title: nberwp 1.1.0
topics: [economics, research]
loadMathJax: no
---

```{r setup, echo = F, message = F, warning = F}
library(bldr)
library(dplyr)
library(ggplot2)
library(knitr)
library(nberwp)

opts_chunk$set(echo = F, message = F, warning = F,
               fig.width = 6, fig.height = 3, fig.ext = 'svg', dev = 'svg')

set_ggtheme()
```

A new version of [nberwp](https://github.com/bldavies/nberwp), an R package containing data on [NBER working papers](https://www.nber.org/papers), is [available on CRAN](https://cran.r-project.org/package=nberwp).
This version adds information about (i) papers published in July--December 2021 and (ii) author sexes.

## Papers from late 2021

```{r new-authors}
# Second half of 2021
new_papers = papers %>%
  filter(year == 2021 & month >= 7)
n_new_paper_authors = new_papers %>%
  left_join(paper_authors) %>%
  {n_distinct(.$author)}
new_authors = authors %>%
  mutate(paper = sub('(.*)[.].*', '\\1', author)) %>%
  semi_join(new_papers)

# Second half of 2020
new_papers_lag = papers %>%
  filter(year == 2020 & month >= 7)
n_new_paper_authors_lag = new_papers_lag %>%
  left_join(paper_authors) %>%
  {n_distinct(.$author)}
new_authors_lag = authors %>%
  mutate(paper = sub('(.*)[.].*', '\\1', author)) %>%
  semi_join(new_papers_lag)
```

The second half of 2021 saw `r nrow(new_papers)` new NBER working papers by `r format(n_new_paper_authors, big.mark = ',')` unique authors, `r nrow(new_authors)` of whom had not published in the series previously.
Those counts were down (from `r nrow(new_papers_lag)`, `r format(n_new_paper_authors_lag, big.mark = ',')`, and `r nrow(new_authors_lag)`, respectively) from the second half of 2020, but roughly in-line with pre-pandemic trends:

```{r monthly-papers}
papers %>%
  filter(year %in% 2017:2021) %>%
  count(Month = as.Date(paste(year, month, 1, sep = '-')), name = 'New papers') %>%
  ggplot(aes(Month, `New papers`)) +
  geom_line() +
  labs(title = 'The pandemic publication boom appears to have ended',
       subtitle = 'New NBER working papers, by month',) +
  coord_cartesian(clip = 'off') +
  scale_x_date(expand = c(0, 0), date_labels = '%b \'%y') +
  scale_y_continuous(expand = c(0, 0))
```

```{r old-authors}
n_old_authors = nrow(authors) - nrow(new_authors)
n_old_authors_prev = 15437
n_old_authors_delta = n_old_authors - n_old_authors_prev
```

nberwp 1.1.0 also corrects some [false merges and splits](/blog/nber-co-authorships/) among authors who published *before* July 2021.
These corrections `r ifelse(n_old_authors_delta > 0, 'raised', 'lowered')` the number of such authors from `r format(n_old_authors_prev, big.mark = ',')` in version 1.0.0 to `r format(n_old_authors, big.mark = ',')` in version 1.1.0.

## Author sexes

nberwp 1.1.0 adds information about author sexes, allowing one to, e.g., visualize the growing [female representation](/blog/female-representation-collaboration-nber/) among NBER working paper authors:

```{r female-representation}
papers %>%
  left_join(paper_authors) %>%
  distinct(year, author) %>%
  left_join(authors) %>%
  group_by(year) %>%
  summarise(p = 100 * mean(female[female >= 0])) %>%
  ggplot(aes(year, p)) +
  geom_line() +
  labs(x = 'Publication year',
       y = '% authors who were female',
       title = 'Growing female representation at the NBER',
       subtitle = 'Share of NBER working paper authors who were female, by publication year') +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))
```

I obtain sex information by matching authors' names with baby name and Facebook data, and through manual identification.
I document my matching and manual procedures in "[Sex-based sorting among economists: Evidence from the NBER](https://doi.org/10.31235/osf.io/zeb7a)," a new paper comparing males' and females' co-authorship patterns.

```{r session-info}
save_session_info()
```
