---
title: nberwp is now on CRAN
topics: [data, research]
loadMathJax: no
---

```{r setup, echo = F, message = F, warning = F}
library(bldr)
library(dplyr)
library(ggplot2)
library(igraph)
library(knitr)
library(nberwp)
library(tidyr)

opts_chunk$set(message = F, warning = F,
               fig.width = 6, fig.height = 3.5, fig.ext = 'svg', dev = 'svg')

set_ggtheme()
```

nberwp, an R package providing information on [NBER working papers](https://www.nber.org/papers) and their authors, is now [available on CRAN](https://cran.r-project.org/package=nberwp).
The current version (1.0.0) covers `r format(nrow(papers), big.mark = ',')` papers published between June 1973 and June 2021.
It can be installed via

```r
install.packages('nberwp')
```

nberwp has evolved since [its initial release](/blog/introducing-nberwp/) on GitHub nearly two years ago.
This post describes some of the main changes.

## More papers

```{r echo = F}
series_counts = papers %>%
  mutate(series = substr(paper, 1, 1)) %>%
  count(series) %>%
  spread(series, n)
```

The first version of nberwp covered papers published between June 1973 and December 2018.
The updated version adds papers published between January 2019 and June 2021, allowing one to visualize the spike in publications when COVID-19 emerged:

```{r covid}
library(dplyr)
library(ggplot2)
library(nberwp)

papers %>%
  count(Quarter = year + (ceiling(month / 3) - 1) / 4, name = 'New papers') %>%
  ggplot(aes(Quarter, `New papers`)) +
  geom_line() +
  labs(title = 'COVID-19 induced a spike in NBER publications',
       subtitle = 'New NBER working papers, by quarter')
```

nberwp now also includes papers published in the historical and technical working paper series.
The historical series contains `r series_counts$h` papers focused on (American) economic history, and the technical series contains `r series_counts$t` papers focused on analytical and empirical methods.

The working paper data exclude duplicates (e.g., papers published in multiple series) but include revisions, which capture continued development of (and collaboration on) research ideas that I believe should be acknowledged.

## Program affiliations

The NBER organizes its research into [programs](https://www.nber.org/programs-projects/programs-working-groups), each of which "corresponds loosely to a traditional field of study within economics."
nberwp now provides a table of paper-program correspondences

```{r}
paper_programs
```

as well as a table of program descriptions:

```{r}
programs
```

```{r echo = F}
paper_program_counts = papers %>%
  left_join(paper_programs) %>%
  group_by(paper) %>%
  summarise(n = sum(!is.na(program)))
```

The `program_category` column categorizes programs similarly to [Chari and Goldsmith-Pinkham (2017)](https://www.nber.org/papers/w23953).
On average, each paper is affiliated with `r round(mean(paper_program_counts$n), 2)` programs and each program has `r format(round(mean(count(paper_programs, program)$n)), big.mark = ',')` affiliated papers.

```{r echo = F}
jaccards = paper_programs %>%
  left_join(paper_programs, by = 'paper') %>%
  count(program.x, program.y) %>%
  group_by(program.x) %>%
  mutate(n.x = sum(n * (program.x == program.y))) %>%
  group_by(program.y) %>%
  mutate(n.y = sum(n * (program.x == program.y))) %>%
  ungroup() %>%
  filter(program.x < program.y) %>%
  mutate(jaccard = n / (n.x + n.y - n)) %>%
  left_join(programs, by = c('program.x' = 'program')) %>%
  left_join(programs, by = c('program.y' = 'program')) %>%
  select(program_desc.x, program_desc.y, jaccard) %>%
  arrange(-jaccard)
```

One use of the paper-program correspondences is to analyze the intellectual overlaps among programs.
For example, the table below presents the six pairs of programs with the most-overlapping sets of affiliated papers, with overlap sizes measured by [Jaccard indices](https://en.wikipedia.org/wiki/Jaccard_index).
The top index of `r round(slice(jaccards, 1)$jaccard, 2)` means that about `r round(100 * slice(jaccards, 1)$jaccard)`% of the papers affiliated with the `r slice(jaccards, 1)$program_desc.x` or `r slice(jaccards, 1)$program_desc.y` programs are affiliated with both.

```{r echo = F}
jaccards %>%
  head() %>%
  kable(digits = 2, col.names = c('Program 1', 'Program 2', 'Jaccard index'))
```

## Authorships

nberwp now contains information about working papers' (co-)authors:

```{r}
authors
```

The `author` column contains unique author identifiers, constructed by concatenating each author's debut paper and their position on that paper's (alphabetized) byline.
This construction ensures that `author` values do not change when I add newly published papers to the data.
The `user_nber` column contains authors' usernames on the NBER website; the `user_repec` column contains authors' [RePEc](https://ideas.repec.org) IDs.
Some authors do not have an NBER username or RePEc ID, indicated by `NA` values in the appropriate column.

nberwp also provides a table of paper-author correspondences:

```{r}
paper_authors
```

```{r echo = F}
net = paper_authors %>%
  left_join(paper_authors, by = 'paper') %>%
  filter(author.x < author.y) %>%
  distinct(author.x, author.y) %>%
  graph_from_data_frame(directed = F, vertices = authors$author)
```

This table can be used to construct a co-authorship network among the `r format(gorder(net), big.mark = ',')` authors identified in nberwp.
This network currently contains `r format(gsize(net), big.mark = ',')` edges, implying that `r round(100 * graph.density(net), 2)`% of pairs co-authored at least one working paper during the period covered by the data.
Authors in the network have a mean degree of `r round(mean(degree(net)), 2)`.

I used previous versions of nberwp in blog posts on [triadic closure](/blog/triadic-closure-nber/) and [female representation](/blog/female-representation-collaboration-nber/).
These posts assumed that authors were uniquely identified by their full names.
This assumption was problematic: different authors could share the same name, or a single author could publish under many names (e.g., before and after marriage).
The updated version of nberwp builds on [previous efforts to disambiguate authors' names](/blog/nber-co-authorships/)---namely cross-referencing against NBER usernames, RePEc IDs, common co-authorships, and name edit distances---in three ways:

1. using paper-program correspondences to identify authors who have similar names and published papers in similar programs, and so are likely to be the same person;
2. manually merging (or splitting) authors whom I determine to be the same (or distinct) based on their personal or academic websites;
3. including an author ID variable (`author`) rather than relying on names for unique identification.

These enhancements support cleaner analyses of (co-)authorship behavior.
Nonetheless the data may still contain errors---if you find any, let me know by adding an issue on [GitHub](https://github.com/bldavies/nberwp).

```{r session-info, echo = F}
save_session_info()
```
