---
title: The option value of waiting
topics: [economics, statistics]
summary: I discuss when and why it can pay to wait for information before taking an action.
---

```{r setup, echo = F, message = F}
library(bldr)
library(dplyr)
library(ggplot2)
library(knitr)
library(purrr)
library(tidyr)

opts_chunk$set(echo = F, message = F,
               fig.width = 6, fig.height = 3, fig.ext = 'svg', dev = 'svg')

set_ggtheme()
```

This post is about waiting for information before taking an action.
It uses a simple model to explain when and why waiting is valuable.
It formalizes some ideas discussed in my posts on [climate change](/blog/climate-change-transport-planning/) and [pandemic](/blog/policymaking-under-uncertainty/) policy.

Suppose it costs $c>0$ to take an action that pays $b>c$ if it is beneficial (`\(\omega=1\)`) and zero otherwise (`\(\omega=0\)`).
I take the action if its expected net benefit
$$\newcommand{\E}{\mathrm{E}}
\E[\omega b-c]=pb-c$$
exceeds zero, where $p=\Pr(\omega=1)$ is my prior belief about $\omega$.
Thus, my decision rule is to take the action whenever $p$ exceeds the cost-benefit ratio $c/b$.

Now suppose I can wait for a [noisy signal](/blog/learning-noisy-signals/) $s\in\{0,1\}$ with error rate
$$\renewcommand{\epsilon}{\varepsilon}
\Pr(s\not=\omega\mid \omega)=\epsilon\in[0,0.5].$$
I use my prior, the signal, and Bayes' rule to form a posterior belief
$$\begin{align}
q_s
&\equiv \Pr(\omega=1\mid s) \\
&= \begin{cases}
\frac{\epsilon p}{(1-\epsilon)(1-p)+\epsilon p} & \text{if}\ s=0 \\
\frac{(1-\epsilon)p}{\epsilon(1-p)+(1-\epsilon)p} & \text{if}\ s=1
\end{cases}
\end{align}$$
about $\omega$.
Then I take the action if its expected net benefit
$$\begin{align}
\E[\omega b-c\mid s]
&= q_sb-c
\end{align}$$
given $s$ exceeds zero.
This happens with probability
$$\Pr(q_sb-c\ge0)=\begin{cases}
1 & \text{if}\ c/b\le q_0 \\
\Pr(s=1) & \text{if}\ q_0<c/b\le q_1 \\
0 & \text{if}\ q_1<c/b,
\end{cases}$$
where the probability
$$\Pr(s=1)=\epsilon(1-p)+(1-\epsilon)p$$
of receiving a positive signal depends on my prior $p$ and the error rate $\epsilon$.

If $c/b\le q_0$ or $q_1<c/b$ then the signal doesn't affect whether I take the action, so I don't need to wait.
But if $q_0<c/b\le q_1$ then waiting gives me a [real option](https://en.wikipedia.org/wiki/Real_options_valuation) not to take the action if I learn it isn't beneficial.
So the expected benefit of waiting equals
$$\begin{align}
W
&\equiv \delta\,\E\left[\E[\max\{0,q_sb-c\}\mid s]\right] \\
&= \begin{cases}
\delta(pb-c) & \text{if}\ c/b\le q_0 \\
\delta(q_1b-c)\Pr(s=1) & \text{if}\ q_0<c/b\le q_1 \\
0 & \text{if}\ q_1<c/b,
\end{cases}
\end{align}$$
where the discount factor $\delta\in[0,1]$ captures
(i) my patience and
(ii) my confidence that the action will still be available if I wait.

I should take the action *before* receiving $s$ if and only if the expected net benefit $(pb-c)$ under my prior exceeds $W$.
This happens precisely when my prior exceeds
$$p^*\equiv\frac{(1-\delta\epsilon)c}{b-\delta((1-\epsilon)b-(1-2\epsilon)c)}.$$
The following chart plots $p^*$ against $\delta$ when $c/b\in\{0.1,0.3,0.5\}$ and $\epsilon\in\{0,0.25,0.5\}$.
Increasing the discount factor $\delta$ or the cost-benefit ratio $c/b$ raises the option value of waiting, which raises the threshold prior $p^*$ above which I should take the action.
Increasing the error rate $\epsilon$ makes the signal less informative, which *lowers* the option value of waiting and, hence, lowers $p^*$.
If $\epsilon=0.5$ then the signal is uninformative and so $p^*=c/b$ independently of $\delta$.

```{r thresholds}
get_threshold = function(alpha, beta, delta, cbr) {
  bcr = 1 / cbr
  N = (1 - delta * alpha)
  D = bcr - delta * ((1 - beta) * bcr - (1 - alpha - beta))
  N / D
}

cbr_values = c(0.5, 0.3, 0.1)
crossing(delta = seq(0, 1, by = 0.01), eps = c(0, 0.25, 0.5), cbr = cbr_values) %>%
  mutate(threshold = pmap_dbl(list(alpha = eps, beta = eps, delta = delta, cbr = cbr), get_threshold),
         cbr = factor(cbr, cbr_values, labels = paste('c/b =', cbr_values))) %>%
  ggplot(aes(delta, threshold, col = cbr)) +
  geom_line() +
  labs(x = 'Discount factor \u03b4',
       y = 'Threshold p*',
       title = 'Threshold priors p* above which I should take the action',
       subtitle = 'p* rises when the discount factor \u03b4 rises, the cost-benefit ratio c/b rises,\nor the error rate \u03b5 falls',
       col = NULL) +
  facet_wrap(~paste('\u03b5 =', eps)) +
  guides(col = guide_legend(label.position = 'left')) +
  coord_cartesian(clip = 'off', expand = F) +
  scale_x_continuous(labels = \(x) round(x, 2)) +
  scale_y_continuous(labels = \(y) round(y, 2), limits = c(0, 1)) +
  scale_color_brewer(palette = 'Set1') +
  theme(legend.justification = c(1, 1),
        legend.position = c(1, 1),
        legend.margin = margin(),
        panel.spacing.x = unit(1, 'line'))
```

```{r session-info}
save_session_info()
```
