---
title: Introducing nberwp
tags: [NBER, research, rstats, text mining]
---

```{r setup, echo = F, message = F}
knitr::opts_chunk$set(message = F, warning = F, paged.print = F,
                      dev = 'svg',
                      fig.ext = 'svg', fig.width = 8, fig.height = 4.5)

library(dplyr)
library(ggplot2)
library(nberwp)

theme_set(
  theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.subtitle = element_text(margin = margin(b = 10), size = 13),
          plot.title = element_text(face = 'bold', margin = margin(b = 10), size = 16),
          strip.text = element_text(face = 'bold', hjust = 0, margin = margin(b = 5), size = 12))
)
```

Today I published [nberwp][nberwp], an R package providing data on [NBER](https://www.nber.org) working papers published between 1973 and 2018.
It can be installed from GitHub via [remotes](https://github.com/r-lib/remotes):

```{r, eval = F}
library(remotes)
install_github('bldavies/nberwp')
```

nberwp provides a data frame `papers`, each row describing a unique working paper:

```{r}
papers
```

`number` uniquely identifies working papers by their positions in the series, while `year` and `month` capture papers' publication dates.
The chart below uses these dates to show the NBER catalogue's expansion.

```{r papers, echo = F}
papers %>%
  count(year) %>%
  mutate(n = cumsum(n)) %>%
  ggplot(aes(year, n / 1e3)) +
  geom_line() +
  labs(x = 'Year',
       y = 'Working papers (000s)',
       title = 'NBER catalogue expansion',
       subtitle = 'Cumulative NBER working paper publications between 1973 and 2018') +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))
```

`title` facilitates simple text mining, such as determining which words are used in working paper titles most frequently:

```{r}
library(tidytext)

words <- papers %>%
  unnest_tokens(word, title) %>%
  anti_join(get_stopwords()) %>%
  filter(nchar(gsub('[a-z.]', '', word)) == 0) %>%
  distinct(number, word)

words %>%
  count(word, sort = T)
```

Many papers discuss capital and labour markets, and the effects of public policies.
The word "evidence" appears in twice as many titles as any other (non-stop) word, which I suspect reflects the growing use of the "\<Issue\>: Evidence from \<context\>" title format:

```{r evidence-from, echo = F}
library(zoo)

papers %>%
  mutate(matches = grepl('evidence from', title, ignore.case = T)) %>%
  group_by(year) %>%
  summarise(share = 100 * mean(matches)) %>%
  mutate(share = rollmean(share, 3, na.pad = T)) %>%
  ggplot(aes(year, share)) +
  geom_line() +
  labs(x = 'Year',
       y = '% of titles',
       title = 'Growing evidence',
       subtitle = 'Share of NBER working paper titles containing \"evidence from\", by year') +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))
```

The NBER's [RePEc index](https://www.nber.org/RePEc/nbr/nberwo/), from which I derive `papers`, also contains data linking papers to their authors.
I plan to include these data in a future version of nberwp once I've disambiguated authors' names.

```{r session-info, echo = F}
options(width = 80)
writeLines(capture.output(sessioninfo::session_info()), 'session.log')
```

[nberwp]: https://github.com/bldavies/nberwp
